### 2017-12-06 manuscript.rmd
minor edits

#### Hypothesis testing
We used the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic] to evaluate two hypotheses: (i) physonects are monophyletic [@totton1965synopsis]; (ii) monoecious species are monophyletic [@Dunn:2005dy]. As the sexual system of *Rudjakovia* sp. is unclear, we carried out two tests of the monophyly of monoecy, one with *Rudjakovia* sp. included as a monoecious species, and one without. We used SOWHAT [@doi:10.1093/sysbio/syv055] dev. version 0.39 (commit fd68ef5733c095c7000a4f92dc8c0daaddeec3b9) to carry out the SOWH tests in parallel with the default options and an initial sample size of 100 (source code can be found in the git repository). For each hypothesis we defined a topology with a single constrained node that was inconsistent with the most likely topology (figure 3 x). We used a threshold for significance of 0.05 and following the initial 100 samples, we evaluated the confidence interval around the p-value to determine if more samples were necessary.

@@

In this phylogeny, *Physophora gilmeri* along with *Lychnagalma utricularia* (both not included in previous phylogeny) are sister to the Agalmatidae *sensu stricto*, a clade restricted to *Agalma*, the Athorybiidae, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *Physophora hydrostatica* (sister species to *P. gilmeri*) was sister to the Forskaliidae with low support. *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of the Agalmatidae *sensu stricto*. Placement of *Cordagalma* outside Agalmatidae *sensu stricto* is consistent with previous analyses using morphological and molecular data [@pugh2006taxonomic;@Dunn:2005dy].

@@

We tested the following three alternative phylogenetic hypotheses against the most likely tree topology: (i) physonect siphonophores are monophyletic, (ii) monoecious siphonophores (not including *Rudjakovia* sp.) are monophyletic, and (iii) monoecious siphonophores (including *Rudjakovia* sp.) are monophyletic (figure 3 xx). In all three cases the alternative hypothesis was rejected (p-value <0.01, confidence interval: <0.001 - 0.03).

@@

Siphonophore colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analyses suggested that the common ancestor of siphonophores was dioecious, and there is a single gain of monoecy within the Codonophora; however the lack of resolution of deep relationships within Codonophora made it impossible to rule out alternative evolutionary scenarios [@Dunn:2005dy]. The current better-resolved tree (Fig. 5a), indicates that monoecy in siphonophores evolved twice from a dioecious ancestor, in the branch leading to Calycophorae and in the branch leading to Agalmatids *sensu lato*. There is a small probability for an alternative scenario featuring a single gain of monoecy before the split of the Codonophora, with a subsequent shift to dioecy in the *Marrus-Erenna* clade.

@@

Nectophores are retained modified medusae that codonophorans use for coordinated colony-level swimming. The nectosome is the region of the colony that develops from the nectosomal growth zone (Fig. 2B). Unlike the siphosomal growth zone, the nectosomal growth zone does not bud gastrozooids, but nectophores (and in the case of *Apolemia*, also palpons). Nectophores are exclusively found on the nectosome -with the exception of  *Physalia physalis* (which has no nectosome, and grows small nectophores near the gonodendra [@Totton1960]. It is possible that the common ancestor of siphonophores had a nectosome, which was lost on the branch leading to cystonects. We cannot exclude with certainty the alternative hypothesis of a nectosome-less ancestor followed by a gain of the nectosome in the branch leading to Codonophora. It is suggested that the nectosome arose as a duplication of the siphosome, followed by functional specialization in propelling the colony [@dunn2006evolution]. The nectosome has been lost within Codonophora in the genus *Athorybia*.

@@

Following the colony development orientation framework [@haddock_dunn_pugh_2005], the nectosome can be located in a dorsal or a ventral position. Our ancestral reconstructions for this character (Fig. S4) show that a ventrally-oriented nectosome was the ancestral form in siphonophores, and that a dorsal nectosome has evolved twice independently, in the braches leading to the Agalmatidae *sensu stricto* and the branch leading to the *Bargmannia* species.

@@

Palpons are modified mouthless gastrozooids used for digestion and circulation of the gastrovascular fluid [@mackie1987siphonophore]. They were present in the common ancestor of siphonophores (Fig. 5b), retained in most species, but lost three times independently in the branches leading to *Pyrostephidae* (represented here by the genera *Bargmannia* and *Rudjakovia*), in calycophorans, and in *Marrus claudanielis*. These taxa might have found other avenues to effectively circulate nutrients across the colony.

@@

The most complex nematocyst batteries of Cnidaria can arguably be found among the siphonophores, hanging in regularly spaced tentacle side branches called tentilla [fig 1A]. Most hydrozoans, including the clade that contains siphonophores, bear simple tentacles (tentacles with no side branches). It is still an open question whether the common ancestor of Siphonophora had simple or branched tentacles. The only siphonophores species regarded as lacking tentilla are *Physalia physalis*, *Apolemia* spp., and *Bathyphysa conifera*. Since *B. conifera* is the only member of the *Rhizophysidae* (and of the *Bathyphysa* genus) lacking tentilla, we can safely assume this is a case of secondary loss. When we reconstruct the evolution of this character on the current phylogeny, 70% of simulations support an common ancestor bearing tentilla, with two independent losses leading to *Physalia* and *Apolemia* (Fig. S2). However, this leaves a 30% support for a simple-tentacled common ancestor followed by 2 independent gains of tentilla in the branches leading to *Rhizophysidae* and *(Bargmannia, Diphyes)*.

How we define absence of tentilla, especially for *Physalia physalis*, is also important. The tentacles of this species, when uncoiled, show very prominent, evenly spaced, bulging buttons which contain on their ectoderm all active and functionally arranged nematocysts used by the organism for prey capture [@Totton1960;@BARDI2007]. Siphonophore tentilla are complete diverticular brachings of the tentacle ectoderm, mesoglea, and gastrovascular canal (lined by endoderm).  *Physalia*'s buttons enclose individual fluid-filled chambers connected by narrow channels to the tentacular canal, lined by endoderm [@Hessinger1988]. This suggests they are not just ectodermal swellings, but probably  reduced tentilla. When we define *Physalia physalis* as tentilla bearing, the results for the character reconstruction lead to a more robust support for a tentilla-bearing common ancestor followed by independent losses of tentilla in the branch leading to *Apolemiidae* (Fig. S3), and in *Bathyphysa conifera*.

@@

Siphonophores are abundant predators in the pelagic realm, ranging from the surface (*Physalia physalis*) to bathypelagic depths (Fig. 4, S5) [@mapstone2014global]. However, the depth distribution of siphonophore populations is not always static. Many species are strong vertical migrators, and some species such as *Chuniphyes multidentata* have exhibited synchronous diel migration patterns [@pugh1984diel]. The phylogeny indicates that the siphonophore common ancestor was probably planktonic, as most extant siphonophore taxa are. Some siphonophore species (such as *Bargmannia lata*) have been observed at abyssopelagic depths near 4000m. Using the present phylogeny, we reconstructed the median depth changes along the phylogeny under a Brownian Motion model (Fig. S5). This model indicates a mesopelagic common ancestor, with several independent transition events to epipelagic (light blue) and bathypelagic (dark blue) waters. This reconstruction only included species recorded using an ROV, thus it excludes many other independent colonizations of the epipelagic habitat. The ROV observations are reliable below 200m, and no quantative measurements were made on SCUBA dives. Species such as *Hippopodius hippopus*, *Athorybia rosacea*, *Diphyes dispar*, and *Chelophyes appendiculata* are often encountered blue water diving less than 20m from the surface (Fig. 4). There was only a single transition to benthic lifestyle on the branch of *Rhodaliidae*, and a single transition to a pleustonic lifestyle on the branch of *Physalia physalis*.

@@

```{r tentilla2, echo=F, message=F, warning=F, fig.width=6.5, fig.height=7, fig.cap= "Stochastic character map of presence of tentilla with Physalia included"}
#Simmap TENTILLA
# plotTree(ultratree, lwd = 4)
# colors = c("black", "red")
# names(colors) = c("Absent", "Present")
# tentilla_sim2 %>% plotSimmap(lwd = 4, add = T,colors=colors)
# nodelabels(pie=(describe.simmap(tentilla_sim2, plot=F)$ace) ,piecol=colors,cex=0.35)
# add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(tentilla_sim2)
```

>>>>

#### Hypothesis testing
We used the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic] to evaluate two hypotheses: (i) physonects are monophyletic [@totton1965synopsis]; (ii) monoecious species are monophyletic [@Dunn:2005dy]. As the sexual system of *Rudjakovia* sp. is unclear, we carried out two tests of the monophyly of monoecy, one with *Rudjakovia* sp. included as a monoecious species, and one without. We used SOWHAT [@doi:10.1093/sysbio/syv055] dev. version 0.39 (commit fd68ef5733c095c7000a4f92dc8c0daaddeec3b9) to carry out the SOWH tests in parallel with the default options and an initial sample size of 100 (source code can be found in the git repository). For each hypothesis we defined a topology with a single constrained node that was inconsistent with the most likely topology (Fig. 3 x). We used a threshold for significance of 0.05 and following the initial 100 samples, we evaluated the confidence interval around the p-value to determine if more samples were necessary.

@@

In this phylogeny, *Physophora gilmeri* along with *Lychnagalma utricularia* (both not included in the previous phylogeny) are sister to the Agalmatidae *sensu stricto*, a clade restricted to *Agalma*, the Athorybiidae, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *Physophora hydrostatica* (sister species to *P. gilmeri*) was sister to the Forskaliidae with low support. *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of the Agalmatidae *sensu stricto*. Placement of *Cordagalma* outside Agalmatidae *sensu stricto* is consistent with previous analyses using morphological and molecular data [@pugh2006taxonomic;@Dunn:2005dy].

@@

We tested the following three alternative phylogenetic hypotheses against the most likely tree topology: (i) physonect siphonophores are monophyletic, (ii) monoecious siphonophores (not including *Rudjakovia* sp.) are monophyletic, and (iii) monoecious siphonophores (including *Rudjakovia* sp.) are monophyletic (Fig. 3 xx). In all three cases the alternative hypothesis was rejected (p-value <0.01, confidence interval: <0.001 - 0.03).

@@

Siphonophore colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analyses suggested that the common ancestor of siphonophores was dioecious, and there is a single gain of monoecy within the Codonophora; however the lack of resolution of deep relationships within Codonophora made it impossible to rule out alternative evolutionary scenarios [@Dunn:2005dy]. The current better-resolved tree (Fig. 5A), indicates that monoecy in siphonophores evolved twice from a dioecious ancestor, in the branch leading to Calycophorae and in the branch leading to Agalmatids *sensu lato*. There is a small probability for an alternative scenario featuring a single gain of monoecy before the split of the Codonophora, with a subsequent shift to dioecy in the *Marrus-Erenna* clade.

@@

Nectophores are retained modified medusae that codonophorans use for coordinated colony-level swimming. The nectosome is the region of the colony that develops from the nectosomal growth zone (Fig. 2B). Unlike the siphosomal growth zone, the nectosomal growth zone does not bud gastrozooids, but nectophores (and in the case of *Apolemia*, also palpons). Nectophores are exclusively found on the nectosome -with the exception of  *Physalia physalis* (which has no nectosome, and grows small nectophores near the gonodendra [@Totton1960]). It is possible that the common ancestor of siphonophores had a nectosome, which was lost on the branch leading to cystonects. We cannot exclude with certainty the alternative hypothesis of a nectosome-less ancestor followed by a gain of the nectosome in the branch leading to Codonophora. It is suggested that the nectosome arose as a duplication of the siphosome, followed by functional specialization in propelling the colony [@dunn2006evolution]. The nectosome has been lost within Codonophora in the genus *Athorybia*.

@@

Following the colony development orientation framework [@haddock_dunn_pugh_2005], the nectosome can be located in a dorsal or a ventral position. Our ancestral reconstructions for this character (Fig. S5) show that a ventrally-oriented nectosome was the ancestral form in siphonophores, and that a dorsal nectosome has evolved twice independently, in the branch leading to the Agalmatidae *sensu stricto* and the branch leading to the *Bargmannia* species.

@@

Palpons are modified mouthless gastrozooids used for digestion and circulation of the gastrovascular fluid [@mackie1987siphonophore]. They were present in the common ancestor of siphonophores (Fig. 5B), retained in most species, but lost three times independently in the branches leading to *Pyrostephidae* (represented here by the genera *Bargmannia* and *Rudjakovia*), in calycophorans, and in *Marrus claudanielis*. These taxa might have found other avenues to effectively circulate nutrients across the colony.

@@

The most complex nematocyst batteries of Cnidaria can arguably be found among the siphonophores, hanging in regularly spaced tentacle side branches called tentilla (Fig. 2A). Most hydrozoans, including the clade that contains siphonophores, bear simple tentacles (tentacles with no side branches). It is still an open question whether the common ancestor of Siphonophora had simple or branched tentacles. The only siphonophores species regarded as lacking tentilla are *Physalia physalis*, *Apolemia* spp., and *Bathyphysa conifera*. Since *B. conifera* is the only member of the *Rhizophysidae* (and of the *Bathyphysa* genus) lacking tentilla, we can safely assume this is a case of secondary loss. When we reconstruct the evolution of this character on the current phylogeny, 70% of simulations support an common ancestor bearing tentilla, with two independent losses leading to *Physalia* and *Apolemia* (Fig. S2). However, this leaves a 30% support for a simple-tentacled common ancestor followed by 2 independent gains of tentilla in the branches leading to *Rhizophysidae* and *(Bargmannia, Diphyes)*.

How we define absence of tentilla, especially for *Physalia physalis*, is also important. The tentacles of this species, when uncoiled, show very prominent, evenly spaced, bulging buttons which contain on their ectoderm all active and functionally arranged nematocysts used by the organism for prey capture [@Totton1960;@BARDI2007]. Siphonophore tentilla are complete diverticular branchings of the tentacle ectoderm, mesoglea, and gastrovascular canal (lined by endoderm).  *Physalia*'s buttons enclose individual fluid-filled chambers connected by narrow channels to the tentacular canal, lined by endoderm [@Hessinger1988]. This suggests they are not just ectodermal swellings, but probably  reduced tentilla. When we define *Physalia physalis* as tentilla bearing, the results for the character reconstruction lead to a more robust support for a tentilla-bearing common ancestor followed by independent losses of tentilla in the branch leading to *Apolemiidae* (Fig. S3), and in *Bathyphysa conifera*.

@@

Siphonophores are abundant predators in the pelagic realm, ranging from the surface (*Physalia physalis*) to bathypelagic depths (Fig. 4, S6) [@mapstone2014global]. However, the depth distribution of siphonophore populations is not always static. Many species are strong vertical migrators, and some species such as *Chuniphyes multidentata* have exhibited synchronous diel migration patterns [@pugh1984diel]. The phylogeny indicates that the siphonophore common ancestor was probably planktonic, as most extant siphonophore taxa are. Some siphonophore species (such as *Bargmannia lata*) have been observed at abyssopelagic depths near 4000m. Using the present phylogeny, we reconstructed the median depth changes along the phylogeny under a Brownian Motion model (Fig. S6). This model indicates a mesopelagic common ancestor, with several independent transition events to epipelagic (light blue) and bathypelagic (dark blue) waters. This reconstruction only included species recorded using an ROV, thus it excludes many other independent colonizations of the epipelagic habitat. The ROV observations are reliable below 200m, and no quantative measurements were made on SCUBA dives. Species such as *Hippopodius hippopus*, *Athorybia rosacea*, *Diphyes dispar*, and *Chelophyes appendiculata* are often encountered blue water diving less than 20m from the surface (Fig. 4). There was only a single transition to benthic lifestyle on the branch of *Rhodaliidae*, and a single transition to a pleustonic lifestyle on the branch of *Physalia physalis*.

@@

```{r tentilla2, echo=F, message=F, warning=F, fig.width=6.5, fig.height=7, fig.cap= "Stochastic character map of presence of tentilla with *Physalia* included"}
#Simmap TENTILLA
# plotTree(ultratree, lwd = 4)
# colors = c("black", "red")
# names(colors) = c("Absent", "Present")
# tentilla_sim2 %>% plotSimmap(lwd = 4, add = T,colors=colors)
# nodelabels(pie=(describe.simmap(tentilla_sim2, plot=F)$ace) ,piecol=colors,cex=0.35)
# add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
densityMap(tentilla_sim2)
```
